---
title: "SMPA - Grouping Assignment"
author: "Kylie Brown"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  During what month were the most disbursements made?

2.  What organization/recepient received the most payments?

3.  What year had the most spending?

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

```{r}

# group by disbursement year and then summarise the total spent - 17082338 in 2017 and 37212716	in 2018
payments %>%
 group_by(disbursement_year) %>%
  summarise(total_spent = sum(disbursement_amount))

```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

```{r}

# group by disbursement month and disbursement year - highest spending month in Jan. 2017
payments %>%
 group_by(disbursement_month, disbursement_year) %>%
  summarise(total_spent = sum(disbursement_amount))

```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

```{r}

#

#filter election cycle to be 2020 and then group by recipient name and create total spent column on totals spent, slice top 10 amount of money received
payments %>% 
  filter(election_cycle == 2020) %>%
  group_by(recipient_name) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 10)
  

```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

```{r}

# filter to top recipient, group by description, summarise total spending and slice descriptions with top 10 highest spending - service was placed media
payments %>% 
  filter(recipient_name == "AMERICAN MADE MEDIA CONSULTANTS, LLC") %>%
  group_by(disbursement_description) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 10)



```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

```{r}

# do same thing as above besides filtering top recipient - most common category was digital consulting/online advertising
payments %>%
  group_by(disbursement_description) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 10)
  

```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

```{r}

# filter disbursement description to legal consulting and then group by recipient name and take sum of spending
payments %>%
  filter(disbursement_description == "LEGAL CONSULTING") %>%
  group_by(recipient_name) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 5)


```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

```{r}
# same thing as above but filter entity type to be for individuals instead of organizations
payments %>%
  filter(entity_type == "IND") %>%
  group_by(recipient_name) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 5)

```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

```{r}

# filter to be for the top individual and group by disbursement and summarise total disbursement amount - top was payroll and 499000.00 spent
payments %>% 
  filter(recipient_name == "GLASSNER, MICHAEL") %>%
  group_by(disbursement_description) %>%
  summarise(total_spent = sum(disbursement_amount)) %>%
  slice_max(total_spent, n = 10)







```
